Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-07 Writing/nw_manuscript.Rnw
manuscript split into two: main and SI

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}
\label{fig:tot-impact}
\end{figure}

@@

%==============================================
%= Supplementary Information
%==============================================
\clearpage

\hypertarget{appendix-appendix}{%
\appendix}

@@

Our analysis evaluates the links between aquifer thickness and drought resilience, proposing a causal link between declining aquifer thickness and irrigated production during droughts. In the first section of our manuscript, we propose that this connection is driven by impacts of aquifer thickness on well yields and, hence, farmers ability to use groundwater to fully buffer crops against water deficits during drought.

@@

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigating farmers in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. As discussed in section \ref{main} of our manuscript, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts.

@@

\caption*{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of corn in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:irrigated-yield-ind-corn}
\end{figure}

%------------------------
%- Figure B.2
%------------------------

@@

\caption*{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of soybean in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:irrigated-yield-ind-soy}
\end{figure}

%------------------------
%- Figure B.3
%------------------------

@@

\caption*{Note: The black solid lines represent average yield of corn (top panel) and soybean (bottom panel) in the US High Plains for different levels of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:avg-yield-ind}
\end{figure}

@@

\caption*{Note: The solid black lines represent the share of acres that are irrigated as a function of aquifer thickness relative to a baseline aquifer thickness level of \Sexpr{sat_ref} m (black line). The gray shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

\setcounter{figure}{0}
\renewcommand{\thefigure}{D.\arabic{figure}}

%------------------------
%- Figure D.1
%------------------------

\caption*{Note: This figure compares the variations in aquifer thickness after demeaning by yearly average (top), state-year average (middle), and county-year average (bottom).}
\label{fig:variation-left}

\end{figure}

%------------------------
%- Figure D.2
%------------------------

\caption*{Note: The solid black lines represent the estimated share of irrigated acres for corn and soybean under different types of fixed effects. Gray shaded area represents 95\% confidence interval.}
\label{fig:state-fe-with-without}
\end{figure}

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{E.\arabic{figure}}

@@

%------------------------
%- Figure E.1
%------------------------

@@

\label{fig:sp-basis}
\end{figure}

@@

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_il_y}

}

\label{fig:ex-fx}
\end{figure}

>>>>

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{\footnotesize{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. Please see section \ref{reg-conf} of the Supplementary Information for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}}
\label{fig:tot-impact}
\end{figure}

@@

%==============================================
%= Supplementary Information
%==============================================
\clearpage
% appendix-appendix

\setcounter{figure}{0}
\renewcommand{\thefigure}{S.\arabic{figure}}
\renewcommand{\thesection}{\Alph{section}}

@@

Our analysis evaluates the links between aquifer thickness and drought resilience, proposing a causal link between declining aquifer thickness and irrigated production during droughts. In the first section of our article, we propose that this connection is driven by impacts of aquifer thickness on well yields and, hence, farmers ability to use groundwater to fully buffer crops against water deficits during drought.

@@

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigating farmers in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. As discussed in section \ref{main} of our article, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts.

@@

\caption*{\footnotesize{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of corn in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}}
\label{fig:irrigated-yield-ind-corn}
\end{figure}

%------------------------
%- Figure S.2
%------------------------

@@

\caption*{\footnotesize{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of soybean in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}}
\label{fig:irrigated-yield-ind-soy}
\end{figure}

%------------------------
%- Figure S.3
%------------------------

@@

\caption*{\footnotesize{Note: The black solid lines represent average yield of corn (top panel) and soybean (bottom panel) in the US High Plains for different levels of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}}
\label{fig:avg-yield-ind}
\end{figure}

@@

\caption*{\footnotesize{Note: The solid black lines represent the share of acres that are irrigated as a function of aquifer thickness relative to a baseline aquifer thickness level of \Sexpr{sat_ref} m (black line). The gray shaded areas represent the 95\% confidence interval for this estimated relationship.}}
\label{fig:test-sat-share}
\end{figure}

@@

@@

%------------------------
%- Figure S.5
%------------------------

@@

\caption*{\footnotesize{Note: This figure presents a series of thin-plate spline basis functions evaluated at a range of $x$ values.}}
\label{fig:sp-basis}
\end{figure}

@@

%------------------------
%- Figure S.6
%------------------------

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_il_y}}
\caption*{\footnotesize{Note: This figure presents two example functions derived by applying different coefficients on the thin-plate spline basis functions.}}
\label{fig:ex-fx}
\end{figure}
\clearpage

%------------------------
%- Figure S.7
%------------------------

\caption*{\footnotesize{Note: This figure compares the variations in aquifer thickness after demeaning by yearly average (top), state-year average (middle), and county-year average (bottom).}}
\label{fig:variation-left}
\end{figure}

%------------------------
%- Figure S.8
%------------------------

\caption*{\footnotesize{Note: The solid black lines represent the estimated share of irrigated acres for corn and soybean under different types of fixed effects. Gray shaded area represents 95\% confidence interval.}}
\label{fig:state-fe-with-without}
\end{figure}
